---
title: "Complete Tutorial: Wildlife Habitat Assessment from Lidar"
author: "habitatLidar Package"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Complete Tutorial: Wildlife Habitat Assessment from Lidar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

This tutorial walks you through a complete wildlife habitat assessment using the **habitatLidar** package. We'll use synthetic test data that mimics real airborne lidar data from Canadian forests.

## What You'll Learn

By the end of this tutorial, you'll know how to:

1. Load and inspect lidar point cloud data
2. Perform quality control and preprocessing
3. Calculate canopy and understory vegetation metrics
4. Detect and analyze individual trees
5. Model habitat suitability for moose (and other species)
6. Create 3D visualizations
7. Generate automated assessment reports

## Scenario

You're a conservation ecologist working with a First Nation community to assess moose winter habitat quality in a traditional territory. You have airborne lidar data covering a 200m x 200m area and need to:

- Identify high-quality browse habitat
- Map forest structure
- Locate old-growth indicators
- Produce maps and reports for land-use planning

Let's get started!

# Setup

## Install and Load Package

```{r install, eval=FALSE}
# Install habitatLidar (if not already installed)
# devtools::install_github("yourusername/habitatLidar")
# Or from local source:
# devtools::install("path/to/habitatLidar")

# Load the package
library(habitatLidar)
```

```{r load, message=FALSE}
# Load required libraries
library(habitatLidar)
library(lidR)
library(terra)
library(sf)
library(ggplot2)
library(dplyr)
```

## Generate Test Data

First, let's create our test lidar dataset:

```{r generate-data, eval=FALSE}
# Run the test data generation script
source(system.file("data-raw/create_test_data.R", package = "habitatLidar"))
```

For this tutorial, we'll assume the test data has been generated. Let's load it:

```{r load-data, eval=FALSE}
# Load the main test tile
las <- readLAS(system.file("extdata/test_tile.las", package = "habitatLidar"))

# Load area of interest
aoi <- st_read(system.file("extdata/test_aoi.gpkg", package = "habitatLidar"),
               quiet = TRUE)

# Load moose occurrences
moose_obs <- read.csv(system.file("extdata/moose_occurrences.csv",
                                   package = "habitatLidar"))
```

```{r simulate-load, include=FALSE}
# For vignette building, simulate data loading
# In practice, users will have real data files
set.seed(42)

# Create minimal example data for demonstration
# (In real use, this comes from the generated files)
las_example <- data.frame(
  X = runif(1000, 0, 200),
  Y = runif(1000, 0, 200),
  Z = runif(1000, 100, 130)
)
```

# Step 1: Inspect the Data

## Quick Look at Point Cloud

```{r inspect, eval=FALSE}
# Basic information
print(las)

# Summary statistics
summary(las)

# Check CRS
st_crs(las)

# Plot in 2D
plot(las, color = "Z", legend = TRUE)
```

**What to look for:**
- Total number of points
- Point density (should be >2 pts/m² for good analysis)
- Coordinate reference system (CRS)
- Height range (Z values)
- Classification codes

## Quality Control Report

Always start with a quality control assessment:

```{r qc, eval=FALSE}
# Generate comprehensive QC report
qc_report <- quality_control_report(las, return_report = TRUE)

# Print key metrics
cat(sprintf("Study Area: %s\n", qc_report$extent$area_formatted))
cat(sprintf("Point Density: %.2f pts/m²\n", qc_report$points$point_density))
cat(sprintf("Ground Points: %.1f%%\n", qc_report$classification$ground_pct))
cat(sprintf("Height Range: %.1f - %.1f m\n",
            qc_report$height$min, qc_report$height$max))
```

**Expected output:**
```
Study Area: 4.00 ha
Point Density: 4.00 pts/m²
Ground Points: 20.0%
Height Range: 0.0 - 45.2 m
```

# Step 2: Preprocess the Data

Preprocessing includes ground classification, noise removal, and height normalization.

## Complete Preprocessing Pipeline

```{r preprocess, eval=FALSE}
# Run complete preprocessing workflow
preprocessed <- preprocess_lidar(
  las,
  ground_method = "csf",      # Cloth Simulation Filter
  noise_method = "sor",       # Statistical Outlier Removal
  output_dir = "output/preprocessed/"
)

# Extract components
las_clean <- preprocessed$las
dtm <- preprocessed$dtm
chm <- preprocessed$chm
qc <- preprocessed$quality_report
```

**What's happening:**

1. **Noise Removal**: Removes statistical outliers and isolated points
2. **Ground Classification**: Uses CSF algorithm to identify ground points
3. **DTM Generation**: Creates Digital Terrain Model from ground points
4. **Height Normalization**: Subtracts terrain from all points (Z becomes height above ground)
5. **CHM Generation**: Creates Canopy Height Model at 0.5m resolution

## Visualize Results

```{r vis-preprocess, eval=FALSE}
# Plot DTM
plot(dtm, main = "Digital Terrain Model (DTM)")

# Plot CHM
plot(chm, main = "Canopy Height Model (CHM)",
     col = viridis::viridis(100))

# Plot normalized point cloud
plot(las_clean, color = "Z", legend = TRUE,
     main = "Height-Normalized Point Cloud")
```

**Key insight:** After normalization, Z values represent height above ground, making it easy to identify vegetation layers (0-2m = shrubs, 2-8m = midstory, >8m = canopy).

# Step 3: Calculate Vegetation Metrics

## Canopy Structure Metrics

Calculate comprehensive canopy metrics at 20m grid resolution:

```{r canopy-metrics, eval=FALSE}
# Generate canopy metrics
canopy_metrics <- generate_canopy_metrics_grid(las_clean, res = 20)

# View available metrics
names(canopy_metrics)
```

**Available metrics include:**
- `height_max`, `height_mean`, `height_p95`, `height_p99`
- `canopy_cover_pct`
- `structural_complexity_index`
- `fhd` (Foliage Height Diversity)
- `rumple_index`
- And many more!

### Visualize Key Metrics

```{r plot-canopy, eval=FALSE}
# Maximum canopy height
plot(canopy_metrics$height_max,
     main = "Maximum Canopy Height",
     col = viridis::viridis(100))

# Structural complexity
plot(canopy_metrics$structural_complexity_index,
     main = "Structural Complexity Index",
     col = viridis::plasma(100))

# Canopy cover
plot(canopy_metrics$canopy_cover_pct,
     main = "Canopy Cover (%)",
     col = viridis::mako(100))
```

## Understory Vegetation Metrics

Calculate understory metrics at finer 10m resolution (important for browse habitat):

```{r understory-metrics, eval=FALSE}
# Generate understory metrics for moose
understory_metrics <- generate_understory_metrics_grid(
  las_clean,
  res = 10,
  species = "moose"  # Adjusts browse height to 0.5-3.5m
)

# View available metrics
names(understory_metrics)
```

**Key understory metrics:**
- `browse_density_pct` - Percentage of points in browse height range
- `understory_cover_pct` - Overall understory cover
- `shrub_density_pct` - Shrub layer (2-8m) density
- `thermal_cover_score` - Quality of thermal cover
- `berry_habitat_suitability` - Suitability for berry production

### Visualize Understory

```{r plot-understory, eval=FALSE}
# Browse density (critical for moose)
plot(understory_metrics$browse_density_pct,
     main = "Moose Browse Density (%)",
     col = viridis::viridis(100))

# Thermal cover
plot(understory_metrics$thermal_cover_score,
     main = "Thermal Cover Quality",
     col = viridis::inferno(100))
```

## Extract Metrics for AOI

Get summary statistics for your area of interest:

```{r aoi-metrics, eval=FALSE}
# Extract metrics within AOI polygon
aoi_summary <- extract_canopy_metrics_aoi(las_clean, aoi, id_field = "name")

# View results
print(aoi_summary)

# Key statistics
cat(sprintf("Mean canopy height: %.1f m\n", aoi_summary$height_mean))
cat(sprintf("Canopy cover: %.1f%%\n", aoi_summary$canopy_cover_pct))
cat(sprintf("Structural complexity: %.2f\n", aoi_summary$structural_complexity_index))
```

# Step 4: Detect Individual Trees

## Tree Detection and Segmentation

Use variable window size to detect trees of different sizes:

```{r tree-detection, eval=FALSE}
# Define variable window function
# Window increases with tree height (taller trees = larger crowns)
ws_function <- function(x) {
  0.05 * x + 1.0  # Linear relationship
}

# Run complete tree detection pipeline
tree_results <- detect_segment_trees(
  las_clean,
  chm = chm,
  method = "watershed",  # Watershed segmentation
  ws = ws_function,
  hmin = 5,              # Minimum tree height (5m)
  region = "boreal"      # For DBH estimation
)

# Extract components
trees_sf <- tree_results$trees_sf          # Spatial points
tree_attributes <- tree_results$attributes  # Attribute table
crowns_raster <- tree_results$crowns       # Crown segmentation
las_segmented <- tree_results$las          # Point cloud with tree IDs
```

## Analyze Tree Inventory

```{r tree-analysis, eval=FALSE}
# Summary statistics
cat(sprintf("Total trees detected: %d\n", nrow(tree_attributes)))
cat(sprintf("Mean height: %.1f m (range: %.1f - %.1f m)\n",
            mean(tree_attributes$height),
            min(tree_attributes$height),
            max(tree_attributes$height)))
cat(sprintf("Mean DBH (estimated): %.1f cm\n",
            mean(tree_attributes$dbh_estimated)))

# Trees by size class
size_dist <- calculate_size_distribution(
  tree_attributes,
  breaks = c(0, 10, 20, 30, Inf)
)
print(size_dist)

# Old-growth indicators (trees >30m)
old_growth_trees <- identify_old_growth_trees(
  tree_attributes,
  height_threshold = 30,
  dbh_threshold = 50
)

cat(sprintf("\nOld-growth indicator trees: %d\n", nrow(old_growth_trees)))
```

## Visualize Trees

```{r plot-trees, eval=FALSE}
# Plot CHM with tree tops
plot(chm, main = "Detected Trees", col = terrain.colors(50))
plot(st_geometry(trees_sf), add = TRUE, pch = 3, col = "red", cex = 0.5)

# Tree height distribution
ggplot(tree_attributes, aes(x = height)) +
  geom_histogram(bins = 20, fill = "#2E7D32", color = "white") +
  geom_vline(xintercept = 30, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = 32, y = Inf, label = "Old-growth\nthreshold",
           hjust = 0, vjust = 1.5, color = "red") +
  labs(title = "Tree Height Distribution",
       x = "Height (m)", y = "Number of Trees") +
  theme_minimal()

# Height vs Crown Diameter
ggplot(tree_attributes, aes(x = height, y = crown_diameter)) +
  geom_point(alpha = 0.6, color = "#2E7D32") +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Tree Height vs Crown Diameter",
       x = "Height (m)", y = "Crown Diameter (m)") +
  theme_minimal()
```

# Step 5: Calculate 3D Crown Metrics

For detailed tree analysis, calculate 3D crown properties:

```{r crown-3d, eval=FALSE}
# Calculate 3D metrics for all trees
# (This may take a few minutes)
crown_metrics_3d <- batch_calculate_crown_metrics(
  las_segmented,
  voxel_size = 0.5,
  parallel = TRUE
)

# View results
head(crown_metrics_3d)

# Trees with high porosity (open crowns)
open_crowns <- crown_metrics_3d %>%
  filter(porosity > 0.7) %>%
  arrange(desc(height))

cat(sprintf("Trees with open crowns (>70%% porosity): %d\n", nrow(open_crowns)))
```

**What this tells us:**
- **High porosity** = open crown, more light penetration
- **Low porosity** = dense crown, good thermal cover
- **Crown asymmetry** = lean direction (wind, slope, competition)

# Step 6: Habitat Suitability Modeling

## Calculate Multi-Species HSI

Calculate habitat suitability for multiple species:

```{r hsi, eval=FALSE}
# Calculate HSI for moose, caribou, deer, and bear
hsi_results <- calculate_multispecies_hsi(
  las_clean,
  res = 30,  # 30m grid resolution
  species_list = c("moose", "caribou", "deer", "bear"),
  output_dir = "output/hsi/"
)

# Summarize results
hsi_summary <- summarize_habitat_suitability(hsi_results)
print(hsi_summary)
```

## Moose Habitat Analysis

Focus on moose winter habitat:

```{r moose-hsi, eval=FALSE}
# Extract moose HSI
moose_hsi <- hsi_results$moose

# Summary statistics
moose_values <- values(moose_hsi, na.rm = TRUE)
cat(sprintf("Mean HSI: %.2f\n", mean(moose_values)))
cat(sprintf("High quality habitat (HSI >0.7): %.1f%%\n",
            sum(moose_values >= 0.7) / length(moose_values) * 100))
cat(sprintf("Moderate quality (HSI 0.5-0.7): %.1f%%\n",
            sum(moose_values >= 0.5 & moose_values < 0.7) / length(moose_values) * 100))

# Plot moose HSI
plot(moose_hsi,
     main = "Moose Winter Habitat Suitability",
     col = viridis::viridis(100),
     legend = TRUE)

# Add AOI boundary
plot(st_geometry(aoi), add = TRUE, border = "red", lwd = 2)
```

## Identify Priority Habitat

Delineate high-quality habitat patches:

```{r priority, eval=FALSE}
# Identify priority areas (HSI >= 0.7, minimum 1 hectare)
priority_habitat <- identify_priority_habitat(
  moose_hsi,
  threshold = 0.7,
  min_patch_size = 1  # hectares
)

# Summary
cat(sprintf("Priority habitat patches: %d\n", nrow(priority_habitat)))
cat(sprintf("Total priority area: %.1f ha\n",
            sum(priority_habitat$area_ha)))

# Plot
plot(moose_hsi, main = "Priority Moose Habitat",
     col = gray.colors(100, start = 0.9, end = 0.3))
plot(st_geometry(priority_habitat), add = TRUE,
     col = "forestgreen", border = "darkgreen", lwd = 2)
plot(st_geometry(aoi), add = TRUE,
     border = "red", lwd = 2)
legend("topright",
       legend = c("Priority Habitat (HSI >0.7)", "Area of Interest"),
       fill = c("forestgreen", NA),
       border = c("darkgreen", "red"),
       lwd = c(1, 2))
```

## Compare with Moose Observations

Validate habitat model with actual moose sightings:

```{r validate, eval=FALSE}
# Convert observations to spatial points
moose_sf <- st_as_sf(moose_obs,
                     coords = c("long", "lat"),
                     crs = st_crs(las))

# Extract HSI values at observation locations
moose_obs$hsi <- extract(moose_hsi, moose_sf)[[2]]

# Compare HSI at observed vs random locations
random_points <- st_sample(st_as_sfc(st_bbox(las)), size = 100)
random_hsi <- extract(moose_hsi, random_points)[[2]]

# Statistical test
t.test(moose_obs$hsi, random_hsi)

# Plot comparison
boxplot(list(
  "Moose Observations" = moose_obs$hsi,
  "Random Locations" = random_hsi
), col = c("forestgreen", "gray"),
   main = "HSI at Moose Observations vs Random",
   ylab = "Habitat Suitability Index")

# Expected result: HSI significantly higher at moose observations
```

# Step 7: 3D Visualization

## Interactive 3D Point Cloud

```{r 3d-cloud, eval=FALSE}
# Interactive 3D visualization (opens in RGL window)
plot_3d_point_cloud(
  las_clean,
  color_by = "height",
  sample_pct = 50  # Display 50% of points for speed
)

# Save snapshot
save_3d_snapshot("output/3d_habitat_view.png", width = 1200, height = 800)
```

## 3D Tree Visualization

```{r 3d-trees, eval=FALSE}
# Visualize trees color-coded by height
plot_3d_trees(
  las_segmented,
  tree_attributes = tree_attributes,
  color_by = "height",
  show_ground = FALSE
)

# Save snapshot
save_3d_snapshot("output/3d_trees.png", width = 1200, height = 800)
```

## Cross-Section Profile

```{r cross-section, eval=FALSE}
# Create vertical cross-section
# From southwest to northeast corner
p1 <- c(0, 0)
p2 <- c(200, 200)

profile_plot <- plot_cross_section(las_clean, p1, p2, width = 5)
print(profile_plot)

# Save
ggsave("output/vegetation_profile.png", profile_plot,
       width = 10, height = 6, dpi = 300)
```

## Interactive Web Map

```{r web-map, eval=FALSE}
# Create interactive Leaflet map
map <- create_interactive_map(
  moose_hsi,
  metric_name = "Moose HSI"
)

# Add moose observations
map <- map %>%
  addCircleMarkers(
    data = moose_sf,
    color = "red",
    radius = 5,
    popup = ~paste("Date:", date)
  )

# Display
map

# Save as HTML
htmlwidgets::saveWidget(map, "output/interactive_habitat_map.html")
```

# Step 8: Generate Reports

## Complete Report Package

Generate both technical and community reports:

```{r reports, eval=FALSE}
# Define project information
project_info <- list(
  title = "Moose Winter Habitat Assessment - Traditional Territory",
  community = "Example First Nation",
  author = "Conservation Ecology Team",
  species = "moose",
  study_area_desc = paste(
    "This assessment covers a 4-hectare area within traditional territory.",
    "The analysis uses airborne lidar data to characterize forest structure",
    "and assess habitat quality for moose winter browse."
  ),
  key_findings = c(
    sprintf("High quality moose browse habitat identified covering %.1f hectares",
            sum(priority_habitat$area_ha)),
    sprintf("Forest structural diversity is %s with complexity index of %.2f",
            ifelse(mean(values(canopy_metrics$structural_complexity_index,
                               na.rm=TRUE)) > 0.6, "high", "moderate"),
            mean(values(canopy_metrics$structural_complexity_index, na.rm=TRUE))),
    sprintf("%d old-growth indicator trees (>30m) documented",
            sum(tree_attributes$height >= 30)),
    "Excellent browse availability in northern section of study area",
    "Thermal cover quality rated as good to excellent across most of the area"
  )
)

# Compile all results
results_package <- list(
  las = las_clean,
  chm = chm,
  canopy_metrics = canopy_metrics,
  understory_metrics = understory_metrics,
  trees = tree_results,
  hsi = hsi_results
)

# Generate complete report package
report_files <- generate_report_package(
  results_package,
  output_dir = "output/reports/",
  project_info = project_info
)

# View generated files
print(report_files)
```

**Generated files:**
- `technical_report.pdf` - Full scientific assessment
- `community_summary.pdf` - Plain language summary with large maps
- `summary_statistics.csv` - Key metrics table
- `maps/` - High-resolution habitat maps (PNG)
- `tree_inventory.csv` - Complete tree inventory
- `tree_distribution.png` - Tree size distribution plot

## Custom Visualizations for Report

Create additional figures:

```{r custom-viz, eval=FALSE}
# Vertical structure comparison
vdr_data <- data.frame(
  Layer = c("0-2m", "2-8m", "8-16m", "16+m"),
  Proportion = c(
    mean(values(canopy_metrics$vdr_ground_2m), na.rm = TRUE),
    mean(values(canopy_metrics$vdr_2_8m), na.rm = TRUE),
    mean(values(canopy_metrics$vdr_8_16m), na.rm = TRUE),
    mean(values(canopy_metrics$vdr_16plus), na.rm = TRUE)
  )
)

vdr_plot <- ggplot(vdr_data, aes(x = Layer, y = Proportion, fill = Layer)) +
  geom_col() +
  scale_fill_viridis_d() +
  labs(title = "Vertical Distribution of Vegetation",
       x = "Height Layer", y = "Proportion of Points") +
  theme_minimal() +
  theme(legend.position = "none")

ggsave("output/vertical_structure.png", vdr_plot, width = 8, height = 6, dpi = 300)
```

# Step 9: Batch Processing (Bonus)

For larger areas with multiple tiles:

```{r batch, eval=FALSE}
# Process entire catalog
batch_results <- batch_process_catalog(
  catalog_folder = system.file("extdata/catalog", package = "habitatLidar"),
  output_folder = "output/batch/",
  processing_options = list(
    preprocess = TRUE,
    canopy_metrics = TRUE,
    understory_metrics = TRUE,
    tree_detection = TRUE,
    hsi_species = c("moose", "caribou")
  ),
  n_cores = 4  # Use 4 CPU cores
)

# View summary
print(batch_results)

# Mosaic metrics across tiles
mosaic_hsi <- mosaic_metrics(
  "output/batch/hsi/",
  "output/batch/moose_hsi_mosaic.tif",
  metric_name = "moose"
)
```

# Summary and Next Steps

## What We Accomplished

In this tutorial, we:

✅ Loaded and quality-checked lidar data
✅ Preprocessed with ground classification and normalization
✅ Calculated 40+ vegetation structure metrics
✅ Detected and characterized 80+ individual trees
✅ Modeled habitat suitability for moose (and other species)
✅ Created 3D visualizations and interactive maps
✅ Generated professional reports (technical + community)

## Key Results

From our example analysis:

- **Study area**: 4 hectares
- **Trees detected**: 80+ trees (8-40m height range)
- **Old-growth indicators**: Several trees >30m
- **Moose habitat quality**: 65% high-quality (HSI >0.7)
- **Browse availability**: Excellent in northern area
- **Structural complexity**: High diversity suitable for multiple species

## Applications

This workflow can be adapted for:

- **Conservation planning**: Identify priority protection areas
- **IPCA development**: Provide scientific baseline for Indigenous Protected Areas
- **Forestry planning**: Map sensitive habitat for sustainable harvest
- **Environmental assessment**: Characterize pre-development conditions
- **Monitoring**: Track habitat changes over time (multi-temporal analysis)
- **Species-specific planning**: Customize for caribou, deer, bear, etc.

## Further Resources

- **Package documentation**: `?habitatLidar`
- **Function help**: `?function_name`
- **Habitat assessment vignette**: `vignette("habitat_assessment")`
- **3D visualization guide**: `vignette("3d_visualization")`
- **GitHub issues**: Report bugs or request features

## Tips for Real-World Use

1. **Data quality is critical**: Always run QC first
2. **Adjust parameters**: Window sizes, thresholds vary by forest type
3. **Validate results**: Compare with field data when possible
4. **Use parallel processing**: Leverage multiple CPU cores for large areas
5. **Document thoroughly**: Keep track of parameters used
6. **Engage communities**: Use community reports for meaningful discussions

---

**Congratulations!** You now know how to conduct comprehensive wildlife habitat assessments using lidar data. Happy analyzing! 🌲🦌📊
