---
title: "Quick Start Guide"
author: "habitatLidar Package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

This quick start guide will get you up and running with **habitatLidar** for analyzing wildlife habitat structure from airborne lidar data. We'll walk through a complete workflow from reading lidar data to generating habitat suitability maps and reports.

## Installation

```{r install}
# Install from GitHub (once published)
# devtools::install_github("yourusername/habitatLidar")

# Or install from local source
# install.packages("habitatLidar", repos = NULL, type = "source")

# Load the package
library(habitatLidar)
```

## Required Dependencies

The package requires several dependencies. Check that they're installed:

```{r check-deps}
required_packages <- c("lidR", "terra", "sf", "rgl", "future", "ggplot2")
check_dependencies(required_packages)
```

# Basic Workflow

## Step 1: Read and Preprocess Lidar Data

```{r read-data}
# Read a single LAS tile
las <- lidR::readLAS("path/to/your/tile.las")

# Or create a catalog for multiple tiles
catalog <- lidR::readLAScatalog("path/to/tiles/folder/")
```

### Quality Control

Before processing, always check data quality:

```{r qc}
# Generate quality control report
quality_control_report(las, return_report = FALSE)  # Prints to console
```

### Preprocessing Pipeline

Run the complete preprocessing workflow:

```{r preprocess}
# Preprocess: ground classification, normalization, CHM generation
preprocessed <- preprocess_lidar(
  las,
  ground_method = "csf",      # Cloth Simulation Filter
  noise_method = "sor",       # Statistical Outlier Removal
  output_dir = "output/preprocessed/"
)

# Extract components
las_clean <- preprocessed$las
chm <- preprocessed$chm
dtm <- preprocessed$dtm
```

## Step 2: Calculate Vegetation Metrics

### Canopy Metrics

```{r canopy-metrics}
# Calculate canopy structure metrics at 20m resolution
canopy_metrics <- generate_canopy_metrics_grid(las_clean, res = 20)

# View available metrics
names(canopy_metrics)

# Plot a metric
plot(canopy_metrics$height_max, main = "Maximum Canopy Height")
```

### Understory Metrics

```{r understory-metrics}
# Calculate understory metrics at 10m resolution
# Specify target species for browse metrics
understory_metrics <- generate_understory_metrics_grid(
  las_clean,
  res = 10,
  species = "moose"
)

# Plot browse density
plot(understory_metrics$browse_density_pct, main = "Moose Browse Density (%)")
```

## Step 3: Detect Individual Trees

```{r tree-detection}
# Define variable window size function for tree detection
# Window increases with tree height
ws_function <- function(x) { 0.05 * x + 1.0 }

# Run complete tree detection pipeline
tree_results <- detect_segment_trees(
  las_clean,
  chm = chm,
  method = "watershed",
  ws = ws_function,
  hmin = 5,              # Minimum tree height (meters)
  region = "boreal"      # For DBH estimation
)

# Extract results
trees_sf <- tree_results$trees_sf          # Spatial features
tree_attributes <- tree_results$attributes  # Attribute table
crowns <- tree_results$crowns              # Crown raster

# View tree statistics
summary(tree_attributes$height)
summary(tree_attributes$dbh_estimated)
```

## Step 4: Calculate Habitat Suitability

```{r hsi}
# Calculate HSI for multiple species
hsi_results <- calculate_multispecies_hsi(
  las_clean,
  res = 30,
  species_list = c("moose", "caribou", "deer", "bear"),
  output_dir = "output/hsi/"
)

# Plot moose HSI
plot(hsi_results$moose,
     main = "Moose Habitat Suitability Index",
     col = viridis::viridis(100))

# Identify priority habitat areas
priority_habitat <- identify_priority_habitat(
  hsi_results$moose,
  threshold = 0.7,           # High quality threshold
  min_patch_size = 1         # Minimum 1 hectare
)

# View priority areas
plot(sf::st_geometry(priority_habitat), col = "red", add = TRUE)
```

## Step 5: Generate Reports

### Complete Processing Workflow

For a single-function complete workflow:

```{r complete-workflow}
# Process everything at once
results <- process_lidar(
  las_file = "path/to/tile.las",
  output_dir = "output/",
  species = "moose",
  metric_res = 20,
  detect_trees = TRUE
)
```

### Generate Reports

```{r reports}
# Define project information
project_info <- list(
  title = "Moose Habitat Assessment - Traditional Territory",
  community = "First Nation Name",
  author = "Conservation Team",
  species = "moose",
  study_area_desc = "500 hectare area in traditional territory",
  key_findings = c(
    "High quality moose browse habitat identified in northern section",
    "Old-growth forest indicators present with 45 trees >30m",
    "Structural diversity supports multiple wildlife species"
  )
)

# Generate complete report package
report_files <- generate_report_package(
  results,
  output_dir = "output/reports/",
  project_info = project_info
)

# Prints paths to technical report, community report, maps, etc.
print(report_files)
```

# Tips and Best Practices

## Data Quality

- Always run quality control before analysis
- Check point density (>2 pts/m² recommended for tree detection)
- Verify coordinate reference system
- Inspect edge artifacts in tiles

## Processing Parameters

### Ground Classification

- **CSF** (Cloth Simulation Filter): Better for complex terrain
- **PMF** (Progressive Morphological Filter): Faster, good for flatter areas

### Tree Detection

- Adjust `hmin` based on minimum tree size of interest
- Use variable window function for mixed-age forests
- For dense young forests, use smaller window: `function(x) { 0.03 * x + 0.5 }`
- For open mature forests, larger window: `function(x) { 0.08 * x + 1.5 }`

### Metric Resolution

- **Canopy metrics**: 20-30m for landscape analysis, 10m for detailed
- **Understory metrics**: 10-20m (finer resolution for browse habitat)
- **HSI**: 30m balances detail and computational efficiency

## Performance Optimization

```{r performance}
# For large areas, use parallel processing
future::plan(future::multisession, workers = 4)

# Process catalog in chunks
lidR::opt_chunk_size(catalog) <- 500  # 500m chunks
lidR::opt_chunk_buffer(catalog) <- 30  # 30m buffer

# Monitor progress
monitor_batch_progress("output/", total_tiles = 100, check_interval = 60)
```

# Next Steps

- **Habitat Assessment**: See `vignette("habitat_assessment")` for detailed species-specific modeling
- **3D Visualization**: See `vignette("3d_visualization")` for creating interactive visualizations
- **Batch Processing**: See documentation for `batch_process_catalog()` for large-area workflows

# Getting Help

```{r help}
# Function documentation
?preprocess_lidar
?calculate_multispecies_hsi

# Package overview
help(package = "habitatLidar")

# Report issues
# https://github.com/yourusername/habitatLidar/issues
```

# Example: Complete Analysis Script

Here's a complete script template:

```{r complete-example}
library(habitatLidar)

# 1. Setup
las_file <- "data/my_tile.las"
output_dir <- "output/"
species <- "moose"

# 2. Process
results <- process_lidar(
  las_file = las_file,
  output_dir = output_dir,
  species = species,
  metric_res = 20,
  detect_trees = TRUE
)

# 3. Analyze
hsi_summary <- summarize_habitat_suitability(results$hsi)
priority_areas <- identify_priority_habitat(
  results$hsi[[species]],
  threshold = 0.7,
  min_patch_size = 1
)

# 4. Report
project_info <- list(
  title = "Wildlife Habitat Assessment",
  community = "Community Name",
  species = species,
  key_findings = c(
    sprintf("%d trees detected", nrow(results$trees$attributes)),
    sprintf("%.1f%% high quality habitat", hsi_summary$high_quality_pct[1])
  )
)

generate_report_package(results, paste0(output_dir, "reports/"), project_info)

# 5. Save key outputs
sf::st_write(priority_areas, paste0(output_dir, "priority_habitat.gpkg"))
write.csv(results$trees$attributes, paste0(output_dir, "tree_inventory.csv"))
```

---

*For more detailed examples, see the other vignettes and package documentation.*
